---
title: 'Survival Analysis: Non Proportional Hazards Model'
author: "Nesma Magdi"
date: "October 21, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Greg)
library(dplyr)
library(tidyr)
library(finalfit)
library(ggplot2)
library(moments)
library(e1071)
library(dummies)
library(dplyr)
library(tidyr)
library(finalfit)
library(dummies)
library(moments)
library(corrplot)
library(scatterplot3d)
library(ggplot2)
library(ggplot2)
library(caret)
library(GGally)
library(ggthemes)
library(broom)
library(dplyr)
library(bindrcpp)
library(caTools)
library(rattle)
library(RColorBrewer)
library(nnet)
library(rpart.plot)
library(dichromat)
library(Rmpfr)
library(digest)
library(convertr)
library(microbenchmark)
library(varhandle)
library(hablar)
library(magrittr)
library(survminer)
library(markovchain)
library(survival)
library(expss)
library(AMR)
library(ggpubr)
library(gplots)
library(devtools)
library(dmetar)
library(grDevices)
library(Hmisc)
library(rattle)
library(reshape)
library(magrittr)
library(ggpubr)
library(gridExtra)
library(PerformanceAnalytics)
library(highcharter)
library(car)
library(coxphw)
library(mgcv)
library(clustcurv)
```

## Ovarian Cancer:

Survival in a randomised trial comparing two treatments for ovarian cancer.
Format

Treatment of patients with advanced ovarian carcinoma (stages IIIB and IV) using either cyclophosphamide alone (1 g/m2) or cyclophosphamide (500 mg/m2) plus adriamycin (40 mg/m2) by iv injection every 3 weeks each produced partial regression in approximately one third of the patients.

```{r ovarian, message=FALSE, warning=FALSE}

data("ovarian")
attach(ovarian)

Status <- ifelse(fustat == 0,"Death", "Censord")
Progression <- ifelse(resid.ds == 1, "No Progression", "Progression")
Treatment <- ifelse(rx == 1, "Endoxan alone", "Endoxan + Adria")
Performance <- ifelse(ecog.ps == 1, "High", "Low")
dummy_Age <- ifelse(age >= 50 ,"Old","Young")

ex <- data.frame(Status, Progression, Treatment, Performance, dummy_Age)

Ovarian <- cbind(ovarian, ex)

Ovarian$dummy_Age <- factor(Ovarian$dummy_Age)
Ovarian$Status<- factor(Ovarian$Status)
Ovarian$Progression <- factor(Ovarian$Progression)
Ovarian$Treatment <- factor(Ovarian$Treatment)
Ovarian$Performance <- factor(Ovarian$Performance)


write.csv(Ovarian, file="ovarian.csv")
ovarian <- read.csv(choose.files(), header = T)
attach(ovarian)

```

## Descrition:

```{r , echo=FALSE}

Abbreviation <- c("futime", "fustat", "age", "resid.ds", "rx", "ecog.ps")

Description <- c("Survival or Censoring time", "Censoring status", "Age in years", "Residual disease present", "Treatment Group", "ECOG Performance Status")

Data <- as.data.frame(Description , Abbreviation)
Data
```

```{r , echo=TRUE}

summary(ovarian)

Count <- table(Status, Treatment)
Percent <- prop.table(Count)*100

persent <- as.data.frame(cbind(Count, Percent))
persent

Tret_Per <- as.data.frame(ovarian %>%
            summary_factorlist("Status", "Treatment", 
                               add_dependent_label = T, p=T, cont_cut = 1))

Tret_Per

```

# Survival Analysis:

*Kaplan-Meier Estimator:

- Pointwise CI S(t):

H0: No difference in Survival between two or more independent groups.

```{r warning=FALSE, , echo=TRUE}

surviv <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
print(surviv, print.rmean = TRUE)

summary(surviv)

ggsurvplot(surviv, data = ovarian, pval = TRUE)

```

The 95% confidence interval of survival time is (464, NA); NA in this case means infinity. A 95% upper confidence limit of NA/infinity is common in survival analysis due to the fact that the data is skewed.

*Comparing Survival Curves:

```{r warning=FALSE, , echo=TRUE}

surviv1 <- survfit(Surv(futime, fustat) ~ Treatment, data = ovarian)
summary(surviv1)

plot(surviv1, ylab = "Survival", xlab = "Time (in days)", 
     col = c("dodgerblue4", "firebrick4"), marktime = TRUE, pval = TRUE)
legend("topright", col = c("dodgerblue4", "firebrick4"), 
       legend = levels(ovarian$Treatment), bty = "n", pch = 19)

```

*Comparing Survival:

```{r echo=TRUE}

surviv2 <- survdiff(Surv(futime, fustat) ~ Treatment, data = ovarian)
surviv2

autoplot(surviv1)

```

H0: rejected (p-value = 0.3).

*Post-hoc analysis:

```{r warning=FALSE, , echo=TRUE}

surviv_post <- pairwise_survdiff(Surv(futime, fustat) ~ Treatment, data = ovarian)

surviv_post

```

=========================================================================

# Cox Proportional Hazards:

-Hazard functions for each individual should be strictily parallel & Hazard ratio is constant over time.

- HR = 1: No effect.
  HR > 1: Increase in Hazards.
  HR < 1: Reduced in Hazards.
  
```{r warning=FALSE, , echo=TRUE}

Model2 <- coxph(Surv(futime, fustat) ~ Treatment+  age + 
                  Progression + Performance, data = ovarian)

summary(Model2)

termplot(Model2, terms = "Treatment")

```

*Interpretation:

1- Estimated Hazard Ratio for Young Ages (<= 50) Versus Old Ages (>50) is 1.133 with confidence interval CI of (1.0335 - 1.242).
Estimated Hazard Ratio for Old Ages Versus Young Ages is 0.8827 with confidence interval CI of (1.0335 - 1.242).

=========================================================================

# Test Significance of variance or interaction term in the Model:

*Wald Test (Maximum Likelihood Estimates):

```{r warning=FALSE, , echo=TRUE}

summary(Model2)$coef

wald <- summary(Model2)$coef[4, 1]/summary(Model2)$coef[4, 3]
pvalue <- wald * pnorm(wald, lower.tail = FALSE)
pvalue
```

*p-value of Treatment, Progression, ECOG Performance status is > 0.05. Fail to Reject Null Hypothesis. 

According to p-value of the test, Fail to Reject null hypothesis (for the Treatment, Progression, ECOG Performance status variable)
Thus, the model must not include this variable.

*Likrlihood Ratio Test (LRT):

LRT Statistics has better statistical properities.
First, The reduced. Second, The full.

```{r warning=FALSE, , echo=TRUE}

Model3 <- coxph(Surv(futime, fustat) ~ age, data = ovarian)

anova(Model3, Model2)

Model2$loglik
```

The First is the log likelihood of a model that contains non of the predictors, So we need the second one.

```{r warning=FALSE, , echo=TRUE}

chi <- 2 * Model2$loglik[2] * Model3$loglik[2]

pvalue2 <- 1 - pchisq(chi, df = 3)

pvalue2

```

In this case, using a=0.05 & testing the significance of Performance variable. We must remove it from the model.

=========================================================================

# Evaluate Proportional Hazard:

*Note:
If the graph of the hazards cross for two or more categories of the predictor of interest: PH assumption not met.

*a- Graphical Approach:

```{r warning=FALSE, , echo=TRUE}

Model4 <- survfit(Surv(futime, fustat) ~ Treatment, data = ovarian)

plot(Model4, fun = "cloglog", xlab = "Time (in days) using log", ylab = "log-log Survival", 
     main = "log-log Curves", col = c("dodgerblue4", "firebrick4"))
legend("topright", col = c("dodgerblue4", "firebrick4"), 
       legend = levels(ovarian$Treatment), bty = "n", pch = 19)

```

It seems that the proportional hazards assumption is violated as the log-log survival curves are not parallel.

*b- Schoenfeld Residuals:

Difference between the observed covariate and the expected given the risk set at that time.

```{r warning=FALSE, , echo=TRUE}

residuals_ov <- cox.zph(Model2)
residuals_ov
par(mfrow = c(1, 2))
plot(residuals_ov, var = 1)
plot(residuals_ov, var = 2)
plot(residuals_ov, var = 3)
plot(residuals_ov, var = 4)

```

It seems that the residuals not flat, not centered about zero..

=========================================================================

# Non Proportional Hazard:

#Stratified Proportional Hazard Model (1- Treatment):

H0: The regression coeffecient are assumed to be the same for each stratum.

Ha: The baseline hazard functions may be different and completely unrelated.

```{r warning=FALSE, , echo=TRUE}

Model5 <- coxph(Surv(futime, fustat) ~ Treatment + strata(dummy_Age), data = ovarian)

summary(Model5)

ggsurvplot(survfit(Model5), data = ovarian, conf.int = T)

```

p-value of Treatment Now (p-value = 0.0429) is Significant, not similar to previous model without stratification (p-value = 0.16158).

```{r warning=FALSE, , echo=TRUE}

Fit2 <- survfit(Surv(futime, fustat) ~ factor(Treatment), data = ovarian)
 autoplot(Fit2)

```

=========================================================================

#Stratified Proportional Hazard Model (2- Progression):

H0: The regression coeffecient are assumed to be the same for each stratum.

Ha: The baseline hazard functions may be different and completely unrelated.

```{r warning=FALSE, , echo=TRUE}

Model5_1 <- coxph(Surv(futime, fustat) ~ Progression + strata(dummy_Age), data = ovarian)

summary(Model5_1)

ggsurvplot(survfit(Model5_1), data = ovarian, conf.int = T)

```

p-value of Treatment Now (p-value = 0.0803) is not Significant, but also not similar to previous model without stratification (p-value = 0.29541).

```{r warning=FALSE, , echo=TRUE}

Fit3 <- survfit(Surv(futime, fustat) ~ factor(Progression), data = ovarian)
 autoplot(Fit3)

```

=========================================================================

#Stratified Proportional Hazard Model (3- Ecog Performance Status):

H0: The regression coeffecient are assumed to be the same for each stratum.

Ha: The baseline hazard functions may be different and completely unrelated.

```{r warning=FALSE, , echo=TRUE}

Model5_2 <- coxph(Surv(futime, fustat) ~ Performance + strata(dummy_Age), data = ovarian)

summary(Model5_2)

ggsurvplot(survfit(Model5_2), data = ovarian, conf.int = T)

```

p-value of Treatment Now (p-value = 0.651) is not Significant, but also not similar to previous model without stratification (p-value = 0.60158).

```{r warning=FALSE, , echo=TRUE}

Fit4 <- survfit(Surv(futime, fustat) ~ factor(Performance), data = ovarian)
 autoplot(Fit4)

```

=========================================================================

#Conclusion:

1- Estimated Hazard Ratio for Old Patients (>= 50) Recieving Cyclophosphamide alone Versus Old Patients (>= 50) Recieving Cyclophosphamide & Adriamycin is 3.463 (Increased Hazards) with confidence interval CI of (1.041 - 11.52).

Estimated Hazard Ratio for Old Patients (>= 50) Recieving Cyclophosphamide & Adriamycin Versus Old Patients (>= 50) Recieving Cyclophosphamide alone is 0.2888 (Reduced Hazards) with confidence interval CI of (1.0335 - 1.242).

2- This means, Recieving Cyclophosphamide (500 mg/m2) plus Adriamycin (40 mg/m2) in old Female Patients (>= 50) Reduce hazard of of death by 28.9%, than Recieving Cyclophosphamide alone (1 g/m2) in old Female Patients (>= 50).

==================================================================================================
